Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-08 ms.tex
MWF comments; UL comment on equilibrium existence

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$: $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_{j=1}^n {M_{i,j}} = 1$ for all \emph{j}); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

% Mean fitness principle
Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}

@@

\begin{proof}
By the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007},
$\bar{w}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.

@@

We turn our attention to the case where transition from a subset of types, $K$, is concurrently controlled.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)}.
\end{aligned}
\end{equation}
Then
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

We havefound a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
Could such an allele increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
Note that $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the The leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\begin{aligned}
\lambda_1=1, \\
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{aligned}
\end{equation}

@@

\begin{proof}
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
\begin{equation}
\bar{w}^* L_{ex} = (I-C+MC) D,
\end{equation}
and since $\bar{w}^*$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium} and \nameref{sec:AppA}),
the leading eigenvalue of $\cl_{ex}$ is $\lambda_1=1$.

Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{D_k}{\bar{w}^*}\Big).
\end{equation}
Evaluating this expression for $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{D_k}{\bar{w}^*}\Big) =  
sign(\bar{w}^* - D_k),
\end{multline}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

>>>>

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$: $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $M_{i,j} \ge 0$ for all $i,j$ and $\sum_{j=1}^n {M_{i,j}} = 1$ for all $j$); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

% Mean fitness principle

Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}
The existence of the leading eigenvalue $\bar{w}^*$ and its corresponding right eigenvector $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix
(the leading eigenvalue $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}).

@@

\begin{proof}
$\bar{w}^*$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.

@@

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
Could such an allele increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eq.~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
Note that, in general, $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

\begin{proof}
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
\begin{equation}
\bar{w}^* \cl_{ex}\big\rvert_{\tilde{C}_k = C_k} = (I-C+MC) D,
\end{equation}
and since $\bar{w}^*$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium} and \nameref{sec:AppA}),
the leading eigenvalue of $\cl_{ex}\big\rvert_{\tilde{C}_k = C_k}$ is $\lambda_1=1$.

Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{D_k}{\bar{w}^*}\Big).
\end{equation}
Evaluating this expression at $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{D_k}{\bar{w}^*}\Big) =  
sign(\bar{w}^* - D_k),
\end{multline}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}
